### 2011-09-26 doc/manuscript.tex
I am focusing on rewriting the manuscript as a resource paper.

%% LyX 1.6.7 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[english]{article}
\usepackage[T1]{fontenc}
\usepackage[latin9]{inputenc}
\usepackage[letterpaper]{geometry}
\geometry{verbose,tmargin=1in,bmargin=1in}
\setlength{\parskip}{\bigskipamount}
\setlength{\parindent}{0pt}
\usepackage{url}
\usepackage{graphicx}
\usepackage{setspace}
\usepackage[authoryear]{natbib}
\usepackage{longtable}
\usepackage{epstopdf}
\doublespacing

@@

Inferring two types of gene transfer using whole genomes of \emph{Streptococcus
pyogenes} and\emph{ S.\ dysgalactiae} subspecies \emph{equisimilis}

@@

\subsubsection*{Version: 0.65 (July 8, 2011) \clearpage{}}

\section*{Abstract}

Using two closely related species of \emph{Streptococcus pyogenes}
(SPY) that is a pervasive human-specific pathogen, and \emph{S.\ dysgalactiae}
ssp.\emph{\ equisimilis} (SDE) that is its closely related species
and an opportunistic human pathogen we investigated two types of gene
transfer: recombining and non-recombining
gene transfer. The former lateral process was studied
with the core genomes of \emph{Streptococcus} genomes, and the latter
process with the pan genomes as well as the core genomes. Two
complementary methods were employed to examine genes and their
fragments that could have been exchanged between SPY and SDE as efforts
of finding their roles in the formation of pathogenicity of the bacteria.
We revisited the problem of directionality of gene transfer between
SPY and SDE, and found evidence supporting that homologous
recombining gene transfer was polarized to be relatively from SPY to SDE based
on the core genomes.  Although we did not find evidence for directional
non-recombining gene transfer, virulence genes appeared to undergo
non-recombining gene transfer more than did non-virulence genes.
Recombining gene transfer was not found to be associated with
virulence genes.  This might indicate that additional genes obtained via
non-recombining gene transfer were more effective than gene fragments
transfer via bacterial recombination for SDE in the acquisition of virulence
genes.  While horizontal gene transfer studies traditionally 
focused on genes as a whole,
we rediscovered smaller gene fragments and their transfer
due to recombination between SPY and SDE in aminoacyl tRNA
synthetase (aaRS), and phosphotransferase systems (PTS).

@@

\textbf{Two lateral processes of gene transfer:} 
Foreign bacterial genes can be laterally
brought into a bacterial genome by two different processes 
\citep{Ochman2001,Lawrence2009}:
homologous recombining transfer and non-recombining transfer.
In the former process donor genes that are homologous to those in
a recipient species replace the indigenous genes. In the latter process
donor genes are integrated into a recipient species without replacement
of genes in the species. 
Although the discrimination between the two
lateral processes was noted a decade ago, methods of detecting horizontal
gene transfer did not appear to properly take the differences into
account. 
Proper discrimination of the two processes of gene transfer
could lead to better understanding of biological nature
of transferred genes. For example, 
bacterial lateral transfers of small fragments of ribosomal RNA genes 
were known to be 
due to homologous recombination \citep{Sneath1993,Yap1999} not non-recombining
gene transfer. 
Because traditionally non-recombining gene transfer in bacteria has been
referred to as horizontal gene transfer, association of ribosomal RNA genes with
just lateral transfer might be misleading.

@@

\textbf{Previous studies on gene transfer and its directionality between
SPY and SDE:} 
Because of the close interaction between SPY and SDE, 
possible donor and recipient relationship of the two species were noted before.
However, previous researches in gene transfer between
the two pyogeneic species of SPY and SDE were limited to small portion
of the bacterial genome, mostly focusing on several genes 
\citep{Kalia2004,Towers2004,Bessen2005}.
For example, \citet{Sachse2002} showed gene transfer from SDE to
SPY for the downstream of the genomic region of streptococcal pyrogenic
exotoxins \emph{speG}. \citet{Davies2007} investigated phage 3396
from SDE as a vehicle of gene transfer from SDE to SPY. \citet{Davies2009}
demonstrated that an integrative conjugative element of SDE allowed
gene transfer from SDE to other streptococci. On the contrary, \citet{Kalia2001}
attempted to show that the direction of gene transfer from SPY to
SDE was more dominant than the other direction, albeit unfortunately
appeared to be premature. Because their researches on direction of
gene flow were limited to only small portion of regions in the genomes,
it remains to be elusive to determine the genome-wide degrees to which
genomic segments flow between the two species and directions of thereof.

@@

\subsection{ClonalOrigin analysis}
%\textbf{Genome sequence alignments and clonal frame:}
The species tree of the five genomes is shown in Figure S\ref{fig:tree5}, which
was estimated using 274 blocks of the core genome alignments with ClonalFrame. 
Figure S\ref{fig:blocksize} shows the distribution of block lengths. The average
length was 4215 bp, and total length of the blocks is 1155016 bp.  There was good 
agreement on the topology among independent chains, and reasonable agreement on 
branch lengths. The tree was also in good agreement with the one used by 
\citet{Suzuki2011}. ClonalFrame estimated the relative strength of recombination
compared to mutations to be $r/m = 4.66$, indicating 
that reconmination-driven substitutions affected more 
sites than did mutation-driven substitutions. The first stage of ClonalOrigin 
was then applied to obtain block-wise estimates of mutation rate, 
recombination rate, and average recombinant tract length 
(Figure \ref{fig:scatter3}).
In the first stage of ClonalOrigin MCMC a few alignment blocks were not
successfully finished even after a month of computation. 
Because those were not expected
to affect the global estimates of population parameters significantly, we used
only finished results of the first stage of ClonalOrigin to estimate mutation
rate, recombination rate, and recombinant tract length. 
The median of mutation
rate per site weighted over the lengths of blocks was estimated to
be 0.081 with (0.067,0.094) of interquartile range (IQR). Mutation rates
per site for many alignment blocks were almost centered around the
gobal median. The median of recombination rate per site was estimated
to be 0.012 with (0.006,0.019) of IQR.
The median value for the recombinant tract length was estimated
to be 744 base pairs with (346,2848) of IQR. 
The estimates of the three parameters were based on one of two independent
MCMC chains. 
The estimates from the second MCMC run were close to those of
the first run: mutation rate estimate of 0.081 with IQR of (0.067,0.099),
recombination rate of 0.012 with IQR of (0.006,0.020), and recombinant tract
length estimate of 723 with IQR of (348,2870).
We considered the similar 
results of the two independent chains as confirmation of 
the convergence of the first stage of ClonalOrigin MCMC.

@@

% Figure 1. A Recombinant Tree of ClonalOrigin Model
\begin{figure}
% \includegraphics[scale=0.42]{figures/mauve-analysis-002}
\includegraphics[scale=0.42]{figures/figure1}
\caption{\label{fig:clonalorigin}Recombinant tree in the ClonalOrigin model.
(A) A recombinant tree relating the five genomes with three recombinant edges.
(B) A genomic region of size 100 base pairs with subregions affected by the
three recombinant edges. Genomic positions are evenly marked by ticks every 
10-th base pair. Local gene trees induced by the recombinant tree are
illustrated for each region covered by some combinations of recombinant edges
along the genomic region (C-G). For example, the local gene tree (E) is created
by all of the three recombinant edges in the subregion between 51 base pairs
and 70 base pairs.}
\end{figure}
\clearpage{}

% Figure 2. Heat maps.
\begin{figure}
\includegraphics[scale=0.45]{figures/figure2}
\caption{\label{fig:Heatmap-of-recombination}Heatmap of recombination events.
Heat map representing the logarithm base 2 of the number of recombination
events inferred relative to its expectation under the prior, for each
donor/recipient pair of branches. The donor is given by the y-axis;
the recipient on the x-axis.  The
green cells were the ones for which the ratios were meaningless because
the recipient species tree branch of the cells was older than the
donor species branch.}
\end{figure}
\clearpage{}%

% Figure 3. The three main parameters.
\begin{figure}
\subfloat[]{\includegraphics[scale=0.5]{figures/scatter-plot-parameter-1-out-theta}}

\subfloat[]{\includegraphics[scale=0.5]{figures/scatter-plot-parameter-1-out-rho}}

\subfloat[]{\includegraphics[scale=0.5]{figures/scatter-plot-parameter-1-out-delta}}
\caption{\label{fig:scatter3}Scatter plots of mutation rate, recombination
rate, and average tract length for the 270 blocks. The red dashed
line is the global median of all the blocks.}
\end{figure}
\clearpage{}%

@@

% Figure 5.
%=============================================================================
% FIGURE - Transfer heatmap
\begin{figure}
\begin{center}
\includegraphics[width=5in]{figures/trans-hgt-heatmap}
\end{center}
\vspace{-1.2in}
\caption{Distribution of horizontal gene transfers.}
\label{fig:hgt-heatmap}
\end{figure}
%=============================================================================

@@

% Figure S4
\begin{figure}
\includegraphics{figures/ri-R}
\caption{\label{fig:ri1}The surrogate measure of recombination intensity
of genes. Each point represents a pair of a true value of average number of
recombinant edges and its estimated value for a gene. The dashed line is with
slope of 1 and intercept of 0.}
\end{figure}
\clearpage{}%

% Figure S5: The species tree
\begin{figure}
\includegraphics[scale=0.7]{figures/species}
\caption{\label{fig:tree5}Species tree of five taxa for ClonalOrigin analysis.
SPY is the branch of a clade\emph{ of Streptococcus pyogenes}, SDE
is that for \emph{S.\ dysgalactiae} ssp.\emph{\ equisimilis}, and SD is the
branch for the three individuals of \emph{S.\ dysgalactiae}. The scale is
proportional to the expected number of mutations given the Watterson's estimate 
of mutation
rate $\theta$ and the tree.}
\end{figure}
\clearpage{}%

% Figure S6: Histogram of block sizes
\begin{figure}
\begin{center}
\includegraphics[width=5in]{figures/blocksize}
\end{center}
\vspace{-.3in}
\caption{Distribution of core alignment block sizes.}
\label{fig:blocksize}
\end{figure}
\clearpage{}%

% Figure S7: gene family sizes
%=============================================================================
% FIGURE - gene family sizes
\begin{figure}
\begin{center}
\includegraphics[width=5in]{figures/famsizes}
\end{center}
\vspace{-.3in}
\caption{Distribution of gene family sizes.}
\label{fig:famsize}
\end{figure}
%=============================================================================

@@

% Figure S8: recombination heatmap from mowgli
%=============================================================================
% FIGURE - Transfer heatmap
\begin{figure}
\begin{center}
\includegraphics[width=5in]{figures/trans-recomb-heatmap}
\end{center}
\vspace{-1.2in}
\caption{Distribution of recombining horizontal gene transfers.}
\label{fig:mowgli-recomb-heatmap}
\end{figure}
%=============================================================================

>>>>

% We approach the project of streptococcus recombination and gene transfer from
% the angle of a resource paper. 
\documentclass[english]{article}
\usepackage[T1]{fontenc}
\usepackage[latin9]{inputenc}
\usepackage[letterpaper]{geometry}
\geometry{verbose,tmargin=1in,bmargin=1in}
\setlength{\parskip}{\bigskipamount}
\setlength{\parindent}{0pt}
\usepackage{url}
\usepackage{graphicx}
\usepackage{setspace}
\usepackage[authoryear]{natbib}
\usepackage{longtable}
\doublespacing

@@

Inferring homologous recombination and horizontal gene transfer 
using whole genomes of \emph{Streptococcus pyogenes} 
and\emph{ S.\ dysgalactiae} subspecies \emph{equisimilis}

@@

\subsubsection*{Version: 0.66 (Sep 30, 2011) \clearpage{}}

\section*{Abstract}
Homologous recombination and horizontal gene transfer are investigated
with complete genomes of closely related bacterial species of 
\emph{Streptococcus pyogenes} (SPY) 
and \emph{S.\ dysgalactiae} subspecies \emph{equisimilis} (SDE).
The former is a pervasive human-specific pathogen, and the latter
an opportunistic human pathogen.  Parts of the genomic configuration, including
human pathogenicity, of SDE could have originated in those of SPY.  
Our genome-wide survey of gene transfer indicated that gene transfer via homologous
recombination occurred relatively from SPY to SDE.
Although we did not find enrichment of virulence genes in the homologous recombining gene
transfer, we found evidence that they transferred more often via
non-recombining fashion than non-virulence genes. 
A genome-wide resource of recombining and non-recombining gene transfer for the study of the
genomes is provided so that research community in pathogenic bacteria can take
advantage of it.

@@

Bacteria evolves new biological features through horizontally transferred genes
from other lineages as well as vertically transmitted mutations on genes within
the same lineage.  Mutations on a gene modifiy relatively small portions of the
gene, including substitutions, insertions, and deletions of deoxyribonucleic
acids. Horizontal gene transfers can reshape the organization of a bacterial
genome, which could revamp the biological function of the bacterial species.
Recipient bacteria can substitute transferred genes for homologous parts of
their genomes. They can also integrate transferred genes in their genomes
without replacement. These two processes are discriminated by calling the former
\textit{recombining gene transfer}, and the latter 
\textit{horizontal gene transfer} \citep{Ochman2001,Lawrence2009}.  
Despite the two distinct evolutionary processes in bacteria, reseraches on
statistical analyses of complete genome sequences in efforts of understanding
bacterial evolution have paid little attention to a combined approach to
illuminating disparity of the two gene transfer processes.

HERE

@@

\textbf{Previous studies on gene transfer and its directionality between
SPY and SDE:} 
Because of the close interaction between SPY and SDE, 
possible donor and recipient relationship of the two species were noted before.
However, previous researches in gene transfer between
the two pyogeneic species of SPY and SDE were limited to small portion
of the bacterial genome, mostly focusing on several genes 
\citep{Kalia2004,Towers2004,Bessen2005}.
For example, \citet{Sachse2002} showed gene transfer from SDE to
SPY for the downstream of the genomic region of streptococcal pyrogenic
exotoxins \emph{speG}. \citet{Davies2007} investigated phage 3396
from SDE as a vehicle of gene transfer from SDE to SPY. \citet{Davies2009}
demonstrated that an integrative conjugative element of SDE allowed
gene transfer from SDE to other streptococci. On the contrary, \citet{Kalia2001}
attempted to show that the direction of gene transfer from SPY to
SDE was more dominant than the other direction, albeit unfortunately
appeared to be premature. Because their researches on direction of
gene flow were limited to only small portion of regions in the genomes,
it remains to be elusive to determine the genome-wide degrees to which
genomic segments flow between the two species and directions of thereof.
Cite this: \citep{Ahmad2009}.

@@

\subsection{ClonalOrigin analysis}
%\textbf{Genome sequence alignments and clonal frame:}
The species tree of the five genomes is shown in Figure S\ref{fig:tree5}, which
was estimated using 274 blocks of the core genome alignments with ClonalFrame. 
Figure S\ref{fig:blocksize} shows the distribution of block lengths. The average
length was 4188 bp.  There was good 
agreement on the topology among independent chains, and reasonable agreement on 
branch lengths. The tree was also in good agreement with the one used by 
\citet{Suzuki2011}. ClonalFrame estimated the relative strength of recombination
compared to mutations to be $r/m = 4.66$, indicating 
that reconmination-driven substitutions affected more 
sites than did mutation-driven substitutions. The first stage of ClonalOrigin 
was then applied to obtain block-wise estimates of mutation rate, 
recombination rate, and average recombinant tract length 
(Figure \ref{fig:scatter3}).
In the first stage of ClonalOrigin MCMC a few alignment blocks were not
successfully finished even after a month of computation. 
Because those were not expected
to affect the global estimates of population parameters significantly, we used
only finished results of the first stage of ClonalOrigin to estimate mutation
rate, recombination rate, and recombinant tract length. 
The median of mutation
rate per site weighted over the lengths of blocks was estimated to
be 0.081 with (0.067,0.094) of interquartile range (IQR). Mutation rates
per site for many alignment blocks were almost centered around the
gobal median. The median of recombination rate per site was estimated
to be 0.012 with (0.006,0.019) of IQR.
The median value for the recombinant tract length was estimated
to be 744 base pairs with (346,2848) of IQR. 
The estimates of the three parameters were based on one of two independent
MCMC chains. 
The estimates from the second MCMC run were close to those of
the first run: mutation rate estimate of 0.081 with IQR of (0.067,0.099),
recombination rate of 0.012 with IQR of (0.006,0.020), and recombinant tract
length estimate of 723 with IQR of (348,2870).
We considered the similar 
results of the two independent chains as confirmation of 
the convergence of the first stage of ClonalOrigin MCMC.

@@

% Figure 1. A Recombinant Tree of ClonalOrigin Model
\begin{figure}
\includegraphics[scale=0.42]{figures/mauve-analysis-002}
\caption{\label{fig:clonalorigin}Recombinant tree in the ClonalOrigin model.
(A) A recombinant tree relating the five genomes with three recombinant edges.
(B) A genomic region of size 100 base pairs with subregions affected by the
three recombinant edges. Genomic positions are evenly marked by ticks every 
10-th base pair. Local gene trees induced by the recombinant tree are
illustrated for each region covered by some combinations of recombinant edges
along the genomic region (C-G). For example, the local gene tree (E) is created
by all of the three recombinant edges in the subregion between 51 base pairs
and 70 base pairs.}
\end{figure}
\clearpage{}

% Figure 2. Heat maps.
\begin{figure}
\includegraphics[scale=0.45]{figures/heatmap-recedge}
\caption{\label{fig:Heatmap-of-recombination}Heatmap of recombination events.
Heat map representing the logarithm base 2 of the number of recombination
events inferred relative to its expectation under the prior, for each
donor/recipient pair of branches. The donor is given by the y-axis;
the recipient on the x-axis.  The
green cells were the ones for which the ratios were meaningless because
the recipient species tree branch of the cells was older than the
donor species branch.}
\end{figure}
\clearpage{}%

% Figure 3. The three main parameters.
\begin{figure}
\subfloat[]{\includegraphics[scale=0.5]{figures/scatter-plot-parameter-1\lyxdot out\lyxdot theta}}

\subfloat[]{\includegraphics[scale=0.5]{figures/scatter-plot-parameter-1\lyxdot out\lyxdot rho}}

\subfloat[]{\includegraphics[scale=0.5]{figures/scatter-plot-parameter-1\lyxdot out\lyxdot delta}}
\caption{\label{fig:scatter3}Scatter plots of mutation rate, recombination
rate, and average tract length for the 270 blocks. The red dashed
line is the global median of all the blocks.}
\end{figure}
\clearpage{}%

@@

% Figure 5.
%=============================================================================
% FIGURE - Transfer heatmap
\begin{figure}
\begin{center}
\includegraphics[width=5in]{figures/trans-hgt-heatmap.pdf}
\end{center}
\vspace{-1.2in}
\caption{Distribution of horizontal gene transfers.}
\label{fig:hgt-heatmap}
\end{figure}
%=============================================================================

@@

% Figure S4
\begin{figure}
\includegraphics{figures/ri1}
\caption{\label{fig:ri1}The surrogate measure of recombination intensity
of genes. Each point represents a pair of a true value of average number of
recombinant edges and its estimated value for a gene. The dashed line is with
slope of 1 and intercept of 0.}
\end{figure}
\clearpage{}%

% Figure S5: The species tree
\begin{figure}
\includegraphics[scale=0.7]{figures/cornellf-3-tree}
\caption{\label{fig:tree5}Species tree of five taxa for ClonalOrigin analysis.
SPY is the branch of a clade\emph{ of Streptococcus pyogenes}, SDE
is that for \emph{S.\ dysgalactiae} ssp.\emph{\ equisimilis}, and SD is the
branch for the three individuals of \emph{S.\ dysgalactiae}. The scale is
proportional to the expected number of mutations given the Watterson's estimate 
of mutation
rate $\theta$ and the tree.}
\end{figure}
\clearpage{}%

% Figure S6: Histogram of block sizes
\begin{figure}
\begin{center}
\includegraphics[width=5in]{figures/blocksize.pdf}
\end{center}
\vspace{-.3in}
\caption{Distribution of core alignment block sizes.}
\label{fig:blocksize}
\end{figure}
\clearpage{}%

% Figure S7: gene family sizes
%=============================================================================
% FIGURE - gene family sizes
\begin{figure}
\begin{center}
\includegraphics[width=5in]{figures/famsizes.pdf}
\end{center}
\vspace{-.3in}
\caption{Distribution of gene family sizes.}
\label{fig:famsize}
\end{figure}
%=============================================================================

@@

% Figure S8: recombination heatmap from mowgli
%=============================================================================
% FIGURE - Transfer heatmap
\begin{figure}
\begin{center}
\includegraphics[width=5in]{figures/trans-recomb-heatmap.pdf}
\end{center}
\vspace{-1.2in}
\caption{Distribution of recombining horizontal gene transfers.}
\label{fig:mowgli-recomb-heatmap}
\end{figure}
%=============================================================================
